---
title: "R24 Interim analysis"
output: html_notebook
---

```{r, echo=FALSE}
#Load packages. You may be asked to update the package 'rlang'
library(ggplot2)
library(tidyverse)
library(phyloseq)
library(vegan)
library(RColorBrewer)
library(patchwork)
library(ggsignif)
library(gridExtra)
```


```{r}
#Load phyloseq object and print it to verify that it was correctly read
ps2 <- read_rds("/Users/jay/Box Sync/R24 16S analysis files/FW__R24_ASV_&_mapping_table/jj_r24_exp_v4tree_noOutliersV3_34T1Added.rds")
ps2
```

```{r}
#Changing variable names for publication
levels(sample_data(ps2)[["Health"]]) <- c("HWC","OB")
ps2
```



Making a plot to show all OTUs on the same scale, pruned for quality

```{r}
#Prune samples that have a total read sum of less than 50

min_count = 50

ps2 %>%
  prune_samples(sample_sums(.)>=min_count, .) ->
  r24_baseline_prune

sample_sums(r24_baseline_prune) %>% sort

```

```{r}
#Only including taxa that have at least 3 reads in at least 10% of samples

min_count = 3
min_sample_frac = 0.10

prune.vec = filter_taxa(r24_baseline_prune, 
                       function(x) sum(x >= min_count) >= (min_sample_frac*length(x)))
r24_baseline.st_prune = prune_taxa(prune.vec, r24_baseline_prune)
ntaxa(r24_baseline.st_prune)

```

```{r}
#Making the dataset "r24_baseline_prune.even" to convert the proportions back into whole numbers (Makes it easier to compare between samples)

r24_baseline_prune.even = transform_sample_counts(r24_baseline.st_prune, function(x) 1E6 * x/sum(x))
```

```{r}
#Plotting even object, sorting by Health
Pruned_Family_by_Health <- plot_bar(r24_baseline_prune.even, x = "Patient", fill = "Family") +
  geom_bar(stat = "identity", position = "stack", size=0) +
  facet_wrap(~Health, scales="free_x")
Pruned_Family_by_Health
```



```{r}
#Transforming ps2 into percentage before pruning
ps2_pct = transform_sample_counts(ps2, function(x) 1E2 * x/sum(x))
```

```{r}
#Pruning transformed ps2 to remove rare taxa to make the legend more readable

#Prune samples that have a total read sum of less than 50

min_count = 50

ps2_pct %>%
  prune_samples(sample_sums(.)>=min_count, .) ->
  ps2_pct_prune

sample_sums(ps2_pct_prune) %>% sort

```

```{r}
#Only including taxa that have at least 3 reads in at least 10% of samples

min_count = 0.01
min_sample_frac = 0.10

prune.vec = filter_taxa(ps2_pct_prune, 
                       function(x) sum(x >= min_count) >= (min_sample_frac*length(x)))
ps2_pct_prune_minreads = prune_taxa(prune.vec, ps2_pct_prune)
ntaxa(ps2_pct_prune_minreads)

```



```{r}
#Making the graph without the trimmed data, to say ASV instead of OTU, and with the Y axis as percentage
Pruned_Family_by_Health31 <- plot_bar(ps2_pct, x = "Patient", fill = "Family") +
  geom_bar(stat = "identity", position = "stack", size=0)  + scale_fill_manual(values = col_vector) + facet_wrap(~Health, scales="free_x") +
  theme(legend.position="bottom") +
  guides(fill=guide_legend(nrow=7))

g7 <- Pruned_Family_by_Health31 + plot_annotation(title = 'Pruned ASVs, sorted by Health') + ylab("Percentage")
  
g7
```

```{r}
#Making the graph without the trimmed data, to say ASV instead of OTU, and with the Y axis as percentage
#This one has a different color palette, see this link for source (https://github.com/joey711/phyloseq/issues/901)
Pruned_Family_by_Health32 <- plot_bar(ps2_pct, x = "Patient", fill = "Family") +
  geom_bar(stat = "identity", position = "stack", size=0)  + scale_fill_manual(values = c("darkblue", "darkgoldenrod1", "darkseagreen", "darkorchid", "darkolivegreen1", "lightskyblue", "darkgreen", "deeppink", "khaki2", "firebrick", "brown1", "darkorange1", "cyan1", "royalblue4", "darksalmon", "darkblue", "royalblue4", "dodgerblue3", "steelblue1", "lightskyblue", "darkseagreen", "darkgoldenrod1", "darkseagreen", "darkorchid", "darkolivegreen1", "brown1", "darkorange1", "cyan1", "darkgrey", "darkblue", "darkgoldenrod1", "darkseagreen", "darkorchid", "darkolivegreen1", "lightskyblue", "darkgreen", "deeppink", "khaki2", "firebrick", "brown1", "darkorange1", "cyan1", "royalblue4", "darksalmon", "darkblue", "royalblue4", "dodgerblue3", "steelblue1", "lightskyblue", "darkseagreen", "darkgoldenrod1", "darkseagreen", "darkorchid", "darkolivegreen1", "brown1", "darkorange1", "cyan1", "darkgrey", "rosybrown4", "darkmagenta", "aquamarine4")) +
theme(legend.position="bottom") + guides(fill=guide_legend(nrow=5)) + facet_wrap(~Health, scales="free_x") +
  theme(legend.position="bottom") +
  guides(fill=guide_legend(nrow=7))

g71 <- Pruned_Family_by_Health32 + plot_annotation(title = 'Pruned ASVs, sorted by Health') + ylab("Percentage")
  
g71
```

```{r}
#Making the graph without the trimmed data, to say ASV instead of OTU, and with the Y axis as percentage
#This is the final version of this plot
Pruned_Family_by_Health32 <- plot_bar(ps2_pct, x = "Patient", fill = "Family") +
  geom_bar(stat = "identity", position = "stack", size=0)  + scale_fill_manual(values = c("burlywood4", "darkgoldenrod1", "darkseagreen", "darkorchid", "darkolivegreen1", "lightskyblue", "darkgreen", "deeppink", "khaki2", "firebrick", "brown1", "darkorange1", "cyan1", "royalblue4", "darksalmon", "darkblue", "limegreen", "dodgerblue3", "tan4", "lightslateblue", "darkseagreen", "darkgoldenrod1", "darkseagreen", "ivory4", "darkolivegreen1", "brown1", "darkorange1", "olivedrab", "darkgrey", "darkblue", "darkgoldenrod4", "darkseagreen", "darkorchid", "darkolivegreen1", "lightskyblue", "darkgreen", "deeppink", "khaki2", "springgreen4", "magenta", "thistle", "slategray2", "royalblue4", "darksalmon", "red2", "khaki4", "slategray2", "plum", "navajowhite3", "sienna2", "darkgoldenrod3", "darkseagreen", "darkorchid", "darkolivegreen1", "brown1", "darkorange1", "cyan1", "darkgrey", "rosybrown4", "darkmagenta", "aquamarine4")) +
theme(legend.position="bottom") + guides(fill=guide_legend(nrow=5)) + facet_wrap(~Health, scales="free_x") +
  theme(legend.position="bottom") +
  guides(fill=guide_legend(ncol=5))

g71 <- Pruned_Family_by_Health32 + ylab("Percentage")
  
g71
```

```{r}
Pruned_Family_by_Health31 <- plot_bar(ps2_pct_prune_minreads, x = "Patient", fill = "Family") +
  geom_bar(stat = "identity", position = "stack", size=0)  + scale_fill_manual(values = col_vector) + facet_wrap(~Health, scales="free_x") +
  theme(legend.position="bottom") +
  guides(fill=guide_legend(nrow=7))

g7 <- Pruned_Family_by_Health31 + plot_annotation(title = 'All OTUs, sorted by Health', subtitle = 'Samples were filtered to include only those that had at least 3 reads in at least 10% of samples, and a total read sum > 50')
  
g7
```


Re-doing boxplots for Alpha and Beta diversity:
Excluded the grey background that is the default for ggplot2
Changed the colors of OB and HWC to match the color scheme in Pat's LefSe analysis

```{r}
#Using Phyloseq and ggplot2 to compare diversity between Health groups (i.e. healthy vs overweight)
#The vector containing diversity metrics can be modified as needed. Here we need only the Shannon diversity index
#This looks good when saved as 4x4
alpha_measures_all <- c("Shannon")
ga1 <- plot_richness(ps2, measures = alpha_measures_all, x ="Health", color = "Health")
ga2 <- ga1 + geom_boxplot(data = ga1$data, aes(x = Health, y = value, color = NULL), alpha = 0.1) + scale_color_brewer(palette = "Set1") + geom_signif(comparisons = list(c("HWC", "OB")), map_signif_level=TRUE)

ga2.1 <- ga2 + geom_jitter(width = 0.3) + theme_minimal()
ga2.1
```

We need to add Kruskal-Wallis statistics to this graph. The GridExtra package is good for this.

```{r}
#Printing alpha beta diversity metrics with a Kruskal wallis test
erich <- estimate_richness(ps2, measures = c("Shannon"))
ttest2 <- t(sapply(erich, function(x) unlist(kruskal.test(x~sample_data(ps2)$Health)[c("estimate","p.value","statistic","conf.int")])))

grid.table(ttest2)
```


Kim's script for Beta Diversity comparisons

```{r}
# normalize counts
jj_r24_even <- rarefy_even_depth(ps2, sample.size=min(sample_sums(ps2)), rngseed=1001, trimOTUs=FALSE)
jj_r24_even <- ps2

HWC_samples <- subset_samples(jj_r24_even, Health=="HWC")
OB_samples <- subset_samples(jj_r24_even, Health=="OB")

# get counts
HWC_counts <- otu_table(HWC_samples)@.Data
no_HWC <- ncol(HWC_counts)
OB_counts <- otu_table(OB_samples)@.Data
no_OB <- ncol(OB_counts)
all_counts <- cbind(HWC_counts, OB_counts)

# note: vegdist assumes samples to compare are listed as rows in this matrix
HWC_within <- vegdist(t(HWC_counts), method="bray", upper=TRUE)
OB_within <- vegdist(t(OBS_counts), method="bray", upper=TRUE)
all_between <- vegdist(t(all_counts), method="bray", upper=TRUE)
HWC_OB_between <- as.matrix(vegdist(t(all_counts), method="bray", upper=TRUE))[1:no_HWC,(no_HWC+1):(no_HWC+no_OBS)]

df <- data.frame(value=c(HWC_within), group="HWC x HWC")
df <- rbind(df, data.frame(value=c(OB_within), group="OB x OB"))
df <- rbind(df, data.frame(value=c(HWC_OB_between), group="HWC x OB"))

p3 <- ggplot(df) +
  geom_boxplot(aes(x=group, y=value, fill=group)) +
  scale_color_brewer(palette="Set1") +
  xlab("") +
  ylab("") +
  theme(legend.position="none")
p3

# test significance of BC dissimilarities
cohorts <- c(rep("HWC", no_HWC), rep("OB", no_OB))
res <- vegan::adonis(all_between ~ cohorts)
R2 <- res$aov.tab$R2
Pval <- res$aov.tab$`Pr(>F)`
cat("PERMANOVA R2 / p-val:",R2[1],"/",Pval[1],"\n")



```

Let's add a title and remove the grey ggplot background

```{r}

p4 <- p3 + ggtitle("Bray-Curtis Dissimilarity comparisons") +
  theme(plot.title = element_text(hjust = 0.5)) + theme_minimal()

p4
```

```{r}
#Using Bray-Curtis dissimilarity with NMDS for ordination
#Uses the pruned dataset generated earlier
#The seed is set at 1 since NMDS can give slightly results each time
set.seed(1)
r24_baseline_prune.even.nmds_bc <- ordinate(r24_baseline_prune.even, "NMDS", "bray")
```


```{r}
#Checking if NMDS converged
cat("Converged?", r24_baseline_prune.even.nmds_bc$converged, fill=TRUE)

```

```{r}
#Checking NMDS stress, this gives the same stress as before which is good
cat("Stress:", r24_baseline_prune.even.nmds_bc$stress, fill=TRUE)

```

###A stress of 0.1-0.2 is good/acceptable. A stress < 0.1 is great

```{r}
## NMDS Scree Plot
#Writing a function to pull stress values from NMDS results for plotting

mds_stress_dplyr = function(df,rep_num, dimensions) {
  mds_result = metaMDS(df, autotransform=TRUE, k=dimensions)
  return(mds_result$stress)
}

```

```{r}
set.seed(1)
scree.df = expand.grid(repnum=seq(1), dimensions=seq(6)) %>% 
  rowwise() %>% 
  mutate(stress = mds_stress_dplyr(otu_table(r24_baseline_prune.even), repnum, dimensions))
```

```{r}
#NMDS plots data using the smallest number of dimensions
ggplot(data = scree.df, aes(x = dimensions, y = stress)) +
  geom_jitter(width = 0.05, alpha=1/3) +
  stat_summary(fun.y=mean, geom="line") +
theme_bw()
```

```{r}
#Plotting the results of NMDS, where each datapoint represents a sample
#This also includes the stress and convergence on the plot
g6 <- plot_ordination(r24_baseline_prune.even, 
                r24_baseline_prune.even.nmds_bc, 
                type="samples", color="Health") +
  scale_color_brewer(palette = "Set1") +
  annotate("text",x=-Inf,y=-Inf,hjust=-0.01,vjust=-0.1,
           label= paste("Stress:", r24_baseline_prune.even.nmds_bc$stress, 
                        "\nConverged:", r24_baseline_prune.even.nmds_bc$converged)) + 
  stat_ellipse(type = "norm")
g6
```

```{r}
#Plotting the results of NMDS, where each datapoint represents a sample
#This also includes the stress and convergence on the plot
g6.1 <- plot_ordination(r24_baseline_prune.even, 
                r24_baseline_prune.even.nmds_bc, 
                type="samples", color="Health", shape = "Gender") +
  scale_color_brewer(palette = "Set1") +
  annotate("text",x=-Inf,y=-Inf,hjust=-0.01,vjust=-0.1,
           label= paste("Stress:", r24_baseline_prune.even.nmds_bc$stress, 
                        "\nConverged:", r24_baseline_prune.even.nmds_bc$converged)) + 
  stat_ellipse(type = "norm")
g6.1
```

```{r}
#Adding 95% confidence ellipses to ordination plots

g7 <- plot_ordination(r24_baseline_prune.even, r24_baseline_prune.even.nmds_bc, type="samples", color="Health") + 
  stat_ellipse(type = "norm") +
  scale_color_brewer(palette = "Dark2") +
  theme_bw()
g7
```

```{r}
#Always provide session info!
sessionInfo()
```

